Visualizing Independence Using Extended Association Plots

### 2004-08-13 trunk/vcd/paper/JCGS.Rnw
started paper for JCGS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@296 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[10pt]{article}
\usepackage{Z}

\setlength{\parskip}{0.5ex plus0.1ex minus0.1ex}
\setlength{\parindent}{0em}

\bibpunct{(}{)}{;}{a}{}{,}

@@

@@

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(12345)
library(vcd)
data(Bundesliga)
goals <- xtabs(~ AwayGoals + HomeGoals, data = Bundesliga, subset = Year == 1995)
data(HairEyeColor)
haireye <- xtabs(Freq ~ Hair + Eye, data = as.data.frame(HairEyeColor), subset = Sex == "Female")
data(UCBAdmissions)
@

\title{Visualizing Independence Using Extended Association and Mosaic Plots}
\author{\hfill Achim Zeileis$^\dag$ \hfill David Meyer$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
\dag {\it \small Institut f\"ur Statistik \& Wahrscheinlichkeitstheorie,
 Technische Universit\"at Wien}\\
\ddag {\it \small Institut f\"ur Statistik, Wirtschaftsuniversit\"at Wien}
}
\date{}
\maketitle

\begin{abstract}
Two visualization techniques for the independence problem in 2-way
contingency tables---assocation and mosaic plots---are extended in two directions:
\begin{enumerate}
\item The visualization is enhanced by displaying the significance of an appropriate
test for independence and by using improved color schemes.
\item The implementation in the \textsf{R} system is improved using a more modular
design and allowing for more flexible specification of plotting parameters.
\end{enumerate}
Furthermore, some enhancements for displays of various types of independence in
general multi-way tables are suggested.
\end{abstract}

@@

\section{Tests for independence in 2-way tables} \label{sec:tests}

@@

\section{Visualization techniques for 2-way tables} \label{sec:visualization}

@@

\subsection{Association plots}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(goals, rot = 0, gp = gp.binary)
@
\caption{Association plot for the Bundesliga data.}
\label{fig:goals-assoc-classic}
\end{center}
\end{figure}

\subsection{Mosaic plots}

@@

\begin{figure}[htbp]
\begin{center}
<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7>>=
mosaicplot(t(goals), shade = TRUE, main = "")
@
\caption{Friendly mosaic plot for the Bundesliga data.}
\label{fig:goals-mosaic-classic}
\end{center}
\end{figure}

@@

\subsection{Visualization enhancements}

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[htbp]
\begin{center}
<<goals-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.assocplot(goals, rot = 0)
@
\caption{Extended association plots for the Bundesliga data.}
\label{fig:goals-assoc-signif}
\end{center}
\end{figure}

To illustrate what a `postive' example looks like we give the association plot
for the well-known Hair-Eye-Color data set (avilable in base \textsf{R}) which cross-tabulates the hair and eye
color and sex of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
shows the association plot for the 328 female students. It is easy to see that
the hypothesis of independence is rejected at 1\% level by the maximum test (\ref{eq:MaxAbs})
due to higher frequencies of students with hair and eye color
black/brown and blond/blue and lower frequencies of
blond/brown and brown/blue than would be plausible under independence. Furthermore,
it can also be seen that the residual corresponding to the black/blue
cell is significant at the 10\% level.

\begin{figure}[htbp]
\begin{center}
<<haireye-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.assocplot(haireye)
@
\caption{Extended association plots for the Hair-Eye-Color data (female students).}
\label{fig:haireye-assoc-signif}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
<<goals-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.mosaicplot(goals, main = NULL, gp = gp.HCLshading)
@
<<haireye-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.mosaicplot(haireye, main = NULL, gp = gp.HCLshading)
@
\caption{Extended mosaic plots for the Bundesliga and Hair-Eye-Color data (female students).}
\label{fig:both-mosaic-signif}
\end{center}
\end{figure}

\subsection{Implementation enhancements}

@@

\section{Multi-way tables}

@@

Two ideas for the problem of conditional independence are briefly outlined
here and illustrated using the famous admissions data of the University of
California at Berkley (UCB) which is available in base \textsf{R}.
In this data, the question whether there is sex
discrimination at the UCB leads to the result that although women seem to
be disadvantaged at the aggregated level there is no sex discrimination
conditioned on the department---with the very exception of one department
in which women are {\em more} likely to be admitted than would be plausible
under independence. Exactly this is illustrated in the conditioning
assocation plot in Figure \ref{fig:UCB-assoc-signif}.

\begin{figure}[htbp]
\begin{center}
<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
tabplot(UCBAdmissions)
@
\caption{Extended conditioned association plots for the UCB Admissions data.}
\label{fig:UCB-assoc-signif}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
##grid.mosaicpairs(UCBAdmissions, type = "conditional", main = NULL)
grid.mosaicpairs(UCBAdmissions, type = "conditional", permute = FALSE, gp = gp.HCLshading)
@
\caption{Extended conditioned association plots for the UCB Admissions data.}
\label{fig:UCB-mosaic-signif}
\end{center}
\end{figure}

\section{Conclusion}
We suggest a set of enhancements for visualizing the independence
problem in 2-way tables with some outlook to multi-way tables. The
extensions aim at improving the visualization by displaying both
the size of the residuals and the significance of a test for
independence and by using better color schemes. Furthermore, a new
implementation is outlined based on the graphics package \texttt{grid}
which provides more modular design and more flexible specification
of graphical parameters.

\bibliographystyle{asa}
\bibliography{vcd}

\end{document}
% LocalWords:  proofness

>>>>

\documentclass[10pt,a4paper]{article}

%% Packages:
\usepackage{graphicx,color,a4wide,thumbpdf,hyperref}

%% \usepackage{Sweave} is essentially
\RequirePackage[T1]{fontenc}
\RequirePackage{ae,fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}

%% BibTeX
\usepackage[authoryear,round,longnamesfirst]{natbib}
\bibliographystyle{asa}
\bibpunct{(}{)}{;}{a}{}{,}

%% new commands
\let\code=\texttt
\let\proglang=\textsf
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\given}{\, | \,}

%% Colors
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}

%% mimic JCGS style
\renewcommand{\section}{\secdef \mysec \mysecnn}
\newcommand{\mysec}[2][default]{\vspace{1.3\baselineskip}%
  \pdfbookmark[1]{#1}{Section.\thesection.#1}%
  \refstepcounter{section}%
  {\begin{center} \large \bf \thesection. \uppercase{#1} \end{center}}%
  \vspace{0.7\baselineskip}}
\newcommand{\mysecnn}[1]{\vspace{1.3\baselineskip}%
  {\begin{center}\large \bf \uppercase{#1} \end{center}}%
  \vspace{0.7\baselineskip}}

\renewcommand{\subsection}{\secdef \mysubsec \mysubsecnn}
\newcommand{\mysubsec}[2][default]{\vspace{1.2\baselineskip}%
  \pdfbookmark[2]{#1}{Subsection.\thesubsection.#1}%
  \refstepcounter{subsection}%
  \noindent%
  {\large \textbf{\thesubsection{} {\sc#1}}} \vspace{.7\baselineskip}}
\newcommand{\mysubsecnn}[1]{\vspace{1.2\baselineskip}%
  \noindent%
  {\large \textbf{{\sc #1}}} \vspace{.7\baselineskip}}
\renewcommand{\abstractname}{}
\renewcommand{\refname}{REFERENCES}

%% Hyperref
\hypersetup{%
  pdftitle = {Visualizing (Conditional) Independence Using Extended Association and Mosaic Plots},
  pdfsubject = {Visualizing (Conditional) Independence},
  pdfkeywords = {contingency tables, association plots, mosaic plots, conditional
    inference, permutation tests},
  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {true},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

\begin{center}
\textbf{\LARGE Visualizing (Conditional) Independence Using Extended Association and Mosaic Plots}

\vspace{0.5cm}
{\large Achim \textsc{Zeileis}, David \textsc{Meyer}, and Kurt \textsc{Hornik}}

\vspace{0.3cm}
\textit{\large Wirtschaftsuniversit\"at Wien, Augasse 2--6, 1090 Wien, Austria}\\[0.14cm]
{\large \textit{E-mail:} \texttt{Firstname.Lastname@wu-wien.ac.at}}
\end{center}

@@

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(12345)
library(vcd)

data(Bundesliga)
goals <- xtabs(~ AwayGoals + HomeGoals, data = Bundesliga, subset = Year == 1995)

data(HairEyeColor)
haireye <- xtabs(Freq ~ Hair + Eye, data = as.data.frame(HairEyeColor), subset = Sex == "Female")

data(UCBAdmissions)
ucb <- UCBAdmissions
dimnames(ucb)[[1]] <- c("Admit", "Reject")

data(Arthritis)
art <- xtabs(~ Improved + Treatment, data = Arthritis)
@

<<auxfunctions,echo=FALSE,results=hide>>=
he.test <- pearson.test(haireye, return = TRUE)
cb <- he.test$qdist(c(0.9, 0.99))

gp.shading <- function(x, y) gp.HSVshading(x, y, lty = 1, test = NULL)
shadeHSVchisq <- function(x, y) gp.HSVshading(x, y, lty = 1)
shadeHSVmax <- function(x, y) gp.HSVshading(x, y, interpolate = cb, lty = 1)
shadeHCLmax <- function(x, y) gp.HCLshading(x, y, interpolate = cb, lty = 1)
shadeHSVmax2 <- function(x, y) gp.HSVshading(x, y, interpolate = c(5,10), test = NULL)
shadeHCLchisq <- function(x, y) gp.HCLshading(x, y, lty = 1)

hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, correct = FALSE)
{
  type <- match.arg(type)
  if(type == "HCL") {
    chroma = seq(0, 100, length = grid.n)
    luminance = seq(0, 100, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, verbose = FALSE, correct = correct))
    xlab <- "chroma"
    ylab <- "luminance"
    main <- paste("hue =", round(hue, digits = 0))
  } else {
    chroma = seq(0, 1, length = grid.n)
    luminance = seq(0, 1, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
    xlab <- "saturation"
    ylab <- "value"
    main <- paste("hue =", round(hue, digits = 3))
  }
  if(plot) {
    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
    for(i in 1:(nc-1)) {
      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
    }
    axis(1)
    axis(2)
    box()
  }
  colnames(color.slice) <- chroma
  rownames(color.slice) <- luminance
  attr(color.slice, "type") <- type
  class(color.slice) <- "slice"
  invisible(color.slice)
}

plot.slice <- function(color.slice)
{
  nc <- ncol(color.slice)
  nl <- nrow(color.slice)
  chroma <- as.numeric(colnames(color.slice))
  luminance <- as.numeric(rownames(color.slice))
  if(attr(color.slice, "type") == "HCL") {
    xlab <- "chroma"
    ylab <- "luminance"
  } else {
    xlab <- "saturation"
    ylab <- "value"
  }
  plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
       xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = deparse(substitute(color.slice)))
  for(i in 1:(nc-1)) {
    rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
  }
  axis(1)
  axis(2)
  box()
}
@

\begin{abstract}
Story starts out not from exploratory analysis with \proglang{R}
but from the problem of visualizing independence. Several
exploratory approaches but not combined with significance
testing.

Paper could be structured in the following way
\begin{itemize}
  \item Introduction
  \item Tests
  \item Data sets
  \item Visualization Techniques
    \begin{itemize}
      \item Mosaic displays
      \item Association plots
      \item Residuals-based shading
    \end{itemize}
  \item Extensions
    \begin{itemize}
      \item max-shading
      \item HCL colors
      \item conditional independence (coplots)
      \item \textit{not the implementation, just mention in summary}
    \end{itemize}
  \item Summary
\end{itemize}

The \proglang{R} side then has to be described in another paper,
probably for the Journal of Statistical Software.\\[0.3cm]
\textbf{Key Words:} Association plots; Conditional inference; 
  Contingency tables; Mosaic plots; Permutation tests.
\end{abstract}

@@

\section{Tests for Independence in 2-way Tables} \label{sec:tests}

@@

\section{Visualization Techniques for 2-way Tables} \label{sec:visualization}

@@

\subsection{Association Plots}

@@

\subsection{Mosaic Plots}

@@

@@

\subsection{Visualization Enhancements}

@@

To illustrate what a `postive' example looks like we give the association plot
for the well-known Hair-Eye-Color data set (avilable in base \textsf{R}) which cross-tabulates the hair and eye
color and gender of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
shows the association plot for the 328 female students. It is easy to see that
the hypothesis of independence is rejected at 1\% level by the maximum test (\ref{eq:MaxAbs})
due to higher frequencies of students with hair and eye color
black/brown and blond/blue and lower frequencies of
blond/brown and brown/blue than would be plausible under independence. Furthermore,
it can also be seen that the residual corresponding to the black/blue
cell is significant at the 10\% level.

@@

\subsection{Implementation Enhancements}

@@

\section{Multi-way Tables}

@@

Two ideas for the problem of conditional independence are briefly outlined
here and illustrated using the famous admissions data of the University of
California at Berkley (UCB) which is available in base \textsf{R}.
In this data, the question whether there is gender
discrimination at the UCB leads to the result that although women seem to
be disadvantaged at the aggregated level there is no gender discrimination
conditioned on the department---with the very exception of one department
in which women are {\em more} likely to be admitted than would be plausible
under independence. Exactly this is illustrated in the conditioning
assocation plot in Figure \ref{fig:UCB-assoc-signif}.

@@

\section{Conclusions}
We suggest a set of enhancements for visualizing the independence
problem in 2-way tables with some outlook to multi-way tables. The
extensions aim at improving the visualization by displaying both
the size of the residuals and the significance of a test for
independence and by using better color schemes. Furthermore, a new
implementation is outlined based on the graphics package \texttt{grid}
which provides more modular design and more flexible specification
of graphical parameters.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(goals, rot = 0, gp = gp.binary)
@
<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaicplot(t(goals), shade = TRUE, main = "")
@
\caption{Classic association and mosaic plot for the Bundesliga data.}
\label{fig:goals-assoc-classic}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<goals-mosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.mosaicplot(goals, gp = gp.max)
@
<<art-mosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.mosaicplot(art, gp = gp.max)
@
\caption{Extended mosaic plots for the Bundesliga and Arthritis data sets.}
\label{fig:assoc}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[htbp]
\begin{center}
<<Friendly-hue,fig=TRUE,echo=FALSE,height=4.8,width=8>>=
plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
dummy <- 0:100/100
dummy.col <- hsv(dummy[-101] + 1/200, 1, 1)
rect(dummy[-101], 0.6, dummy[-1], 1, border = dummy.col, col = dummy.col)
rect(0, 0.6, 1, 1)
axis(3)
mtext("hue", line = 2, cex = 1.5)
text(1, 0.45, "saturation = 1\nvalue = 1", pos = 2, cex = 1.2)
lines(c(0.1, 0), c(0.3, 0.6), lty = 2)
lines(c(2/3, 2/3), c(0.3, 0.6), lty = 2)
rect(0, 0.1, 0.2, 0.3, col = hsv(0, 1, 1))
rect(2/3 - 0.1, 0.1, 2/3 + 0.1, 0.3, col = hsv(2/3, 1, 1))
text(0.1, 0.1, expression(r[ij] < 0), pos = 1, cex = 1.5)
text(2/3, 0.1, expression(r[ij] > 0), pos = 1, cex = 1.5)
@

<<Friendly-sat,fig=TRUE,echo=FALSE,height=4.8,width=8>>=
plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
dummy <- 0:100/100
dummy.col <- hsv(2/3, dummy[-101] + 1/200, 1)
rect(dummy[-101], 0.8, dummy[-1], 1, border = dummy.col, col = dummy.col)
dummy.col <- hsv(0, dummy[-101] + 1/200, 1)
rect(dummy[-101], 0.6, dummy[-1], 0.8, border = dummy.col, col = dummy.col)
rect(0, 0.8, 1, 1)
rect(0, 0.6, 1, 0.8)
axis(3)
mtext("saturation", line = 2, cex = 1.5)
mtext(c("h = 0", "h = 2/3"), side = 2, at = c(0.7, 0.9), las = 1, cex = 1.2)
text(0.8, 0.5, "value = 1", pos = 2, cex = 1.2)
lines(c(0.1, 0), c(0.3, 0.6), lty = 2)
lines(c(0.5, 0.5), c(0.3, 0.6), lty = 2)
lines(c(0.9, 1), c(0.3, 0.6), lty = 2)
rect(0, 0.2, 0.2, 0.3, col = hsv(2/3, 0, 1))
rect(0, 0.1, 0.2, 0.2, col = hsv(0, 0, 1))
rect(0.4, 0.2, 0.6, 0.3, col = hsv(2/3, 0.5, 1))
rect(0.4, 0.1, 0.6, 0.2, col = hsv(0, 0.5, 1))
rect(0.8, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
rect(0.8, 0.1, 1, 0.2, col = hsv(0, 1, 1))
text(0.1, 0.1, expression(abs(r[ij]) < 2), pos = 1, cex = 1.5)
text(0.5, 0.1, expression(2 < abs(r[ij]) < 4), pos = 1, cex = 1.5)
text(0.9, 0.1, expression(abs(r[ij]) > 4), pos = 1, cex = 1.5)
@

<<Friendly-val,fig=TRUE,echo=FALSE,height=4.8,width=8>>=
plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
dummy <- 0:100/100
dummy.col <- hsv(2/3, 1, dummy[-101] + 1/200)
rect(dummy[-101], 0.8, dummy[-1], 1, border = dummy.col, col = dummy.col)
dummy.col <- hsv(0, 1, dummy[-101] + 1/200)
rect(dummy[-101], 0.6, dummy[-1], 0.8, border = dummy.col, col = dummy.col)
rect(0, 0.8, 1, 1)
rect(0, 0.6, 1, 0.8)
axis(3)
mtext("value", line = 2, cex = 1.5)
mtext(c("h = 0", "h = 2/3"), side = 2, at = c(0.7, 0.9), las = 1, cex = 1.2)
text(0, 0.5, paste("saturation =", 1), pos = 4, cex = 1.2)
lines(c(0.85, 1), c(0.3, 0.6), lty = 2)
rect(0.7, 0.2, 0.8, 0.3, col = hsv(2/3, 0, 1))
rect(0.8, 0.2, 0.9, 0.3, col = hsv(2/3, 0.5, 1))
rect(0.9, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
rect(0.7, 0.1, 0.8, 0.2, col = hsv(0, 0, 1))
rect(0.8, 0.1, 0.9, 0.2, col = hsv(0, 0.5, 1))
rect(0.9, 0.1, 1, 0.2, col = hsv(0, 1, 1))
lines(c(0.5, 0.5), c(0.3, 0.6), lty = 2)
lines(c(0.85, 1), c(0.3, 0.6), lty = 2)
rect(0.35, 0.2, 0.45, 0.3, col = hsv(2/3, 0, 0.5))
rect(0.45, 0.2, 0.55, 0.3, col = hsv(2/3, 0.5, 0.5))
rect(0.55, 0.2, 0.65, 0.3, col = hsv(2/3, 1, 0.5))
rect(0.35, 0.1, 0.45, 0.2, col = hsv(0, 0, 0.5))
rect(0.45, 0.1, 0.55, 0.2, col = hsv(0, 0.5, 0.5))
rect(0.55, 0.1, 0.65, 0.2, col = hsv(0, 1, 0.5))
rect(0.7, 0.2, 0.8, 0.3, col = hsv(2/3, 0, 1))
rect(0.8, 0.2, 0.9, 0.3, col = hsv(2/3, 0.5, 1))
rect(0.9, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
rect(0.7, 0.1, 0.8, 0.2, col = hsv(0, 0, 1))
rect(0.8, 0.1, 0.9, 0.2, col = hsv(0, 0.5, 1))
rect(0.9, 0.1, 1, 0.2, col = hsv(0, 1, 1))
text(0.5, 0.1, "non-significant", pos = 1, cex = 1.5)
text(0.85, 0.1, "significant", pos = 1, cex = 1.5)
@

\caption{Extended Friendly shading in HSV space.}
\label{fig:Friendly-HSV}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<Friendly-HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
mysaturation <- as.numeric(colnames(hue23))
value <- as.numeric(rownames(hue23))

## select those with value >= 0.5
hue23 <- hue23[value >= .5, ]
hue0 <- hue0[value >= .5, ]
value <- value[value >= .5]
nl <- nrow(hue23)
nc <- ncol(hue23)

## plot 2 slides from HSV space
plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
for(i in 1:(nc-1)) {
  rect(mysaturation[i], value[-nl], mysaturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
}
for(i in 1:(nc-1)) {
  rect(-mysaturation[i], value[-nl], -(mysaturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
}
axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
mtext("saturation", side = 3, at = 0, line = 2)
mtext("value", side = 2, at = .75, line = 2)
mtext("value", side = 4, at = .75, line = 2)
lines(c(-1, 1), c(.5, .5))

## significant colors
rect(-1, 0.95, -.90, 1, col = hsv(0, 1, 1))
rect(-0.45, 0.95, -.55, 1, col = hsv(0, 0.5, 1))
rect(-.05, .95, .05, 1, col = hsv(2/3, 0, 1))
rect(0.45, 0.95, .55, 1, col = hsv(2/3, 0.5, 1))
rect(.90, .95, 1, 1, col = hsv(2/3, 1, 1))

text(-1, .33, "significant", pos = 4, cex = 1.2)
rect(-1, .20, -.80, .30, col = hsv(0, 1, 1))
rect(-.40, .20, -0.6, .30, col = hsv(0, 0.5, 1))
rect(-.20, .20, 0, .30, col = hsv(0, 0, 1))
rect(0, .20, .20, .30, col = hsv(2/3, 0, 1))
rect(0.4, .20, .60, .30, col = hsv(2/3, .5, 1))
rect(.80, .20, 1, .30, col = hsv(2/3, 1, 1))

lines(c(-.9, -.55), c(0.975, .975), lty = 2)
lines(c(-.45, -.05), c(0.975, .975), lty = 2)
lines(c(.45, .05), c(0.975, .975), lty = 2)
lines(c(.9, .55), c(0.975, .975), lty = 2)

## non-significant colors
rect(-1, 0.5, -.90, 0.55, col = hsv(0, 1, 0.5))
rect(-0.45, 0.5, -.55, 0.55, col = hsv(0, 0.5, 0.5))
rect(-.05, .5, .05, 0.55, col = hsv(2/3, 0, 0.5))
rect(0.45, 0.5, .55, 0.55, col = hsv(2/3, 0.5, 0.5))
rect(.90, .5, 1, 0.55, col = hsv(2/3, 1, 0.5))

text(-1, .13, "non-significant", pos = 4, cex = 1.2)
rect(-1, 0, -.80, .10, col = hsv(0, 1, 0.5))
rect(-.60, 0, -.4, .10, col = hsv(0, 0.5, 0.5))
rect(-.20, 0, 0, .10, col = hsv(0, 0, 0.5))
rect(0, 0, .20, .10, col = hsv(2/3, 0, 0.5))
rect(0.4, 0, .60, .1, col = hsv(2/3, .5, 0.5))
rect(.80, 0, 1, .10, col = hsv(2/3, 1, 0.5))

lines(c(-.9, -.55), c(0.525, .525), lty = 2)
lines(c(-.45, -.05), c(0.525, .525), lty = 2)
lines(c(.45, .05), c(0.525, .525), lty = 2)
lines(c(.9, .55), c(0.525, .525), lty = 2)
@

<<HCL-space,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
mychroma <- as.numeric(colnames(hue260))
luminance <- as.numeric(rownames(hue260))

## select those with lumincance >= 50
hue260 <- hue260[luminance >= 50, ]
hue360 <- hue360[luminance >= 50, ]
luminance <- luminance[luminance >= 50]
nc <- ncol(hue260)
nl <- nrow(hue260)

## plot 2 slides from HCL space
plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
for(i in 1:(nc-1)) {
  rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
}
for(i in 1:(nc-1)) {
  rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
}
axis(2, at = c(50, 75, 100), labels = c(50, 75, 100))
axis(4, at = c(50, 75, 100), labels = c(50, 75, 100))
axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
mtext("chroma", side = 3, at = 0, line = 2)
mtext("luminance", side = 2, at = 75, line = 2)
mtext("luminance", side = 4, at = 75, line = 2)
lines(c(-100, 100), c(50, 50))

## significant colors
rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))

text(-100, 33, "significant", pos = 4, cex = 1.2)
rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
rect(0, 20, 20, 30, col = hcl(260, 0, 90))
#white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
#white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
rect(40, 20, 60, 30, col = hcl(260, 50, 70))
rect(80, 20, 100, 30, col = hcl(260, 100, 50))

lines(c(-45, -5), c(72.5, 87.5), lty = 2)
lines(c(45, 5), c(72.5, 87.5), lty = 2)
lines(c(-95, -55), c(52.5, 67.5), lty = 2)
lines(c(95, 55), c(52.5, 67.5), lty = 2)

## non-significant colors
rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))

text(-100, 13, "non-significant", pos = 4, cex = 1.2)
rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
rect(0, 0, 20, 10, col = hcl(260, 0, 90))
rect(20, 0, 40, 10, col = hcl(260, 10, 70))
rect(40, 0, 60, 10, col = hcl(260, 20, 50))

lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
@
\caption{Extended shading in HSV and HCL space.}
\label{fig:ExtHSVHCL}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic1,fig=TRUE,echo=FALSE,height=4,width=8>>=
fm <- loglm(~(Gender + Admit)*Dept, data = UCBAdmissions, fitted = TRUE)
res.fm <- (UCBAdmissions - fitted(fm))/sqrt(fitted(fm))
dim(res.fm) <- dim(UCBAdmissions)
my.gp <- gp.HCLshading(UCBAdmissions, aperm(res.fm))
my.gp$p.value <- summary(fm)$tests[2,3]
my.gp$lty <- c(1,1)
grid.mosaicplot(ucb, gp = my.gp, margin = ~ (Gender + Admit)*Dept, space = c(0.1, 0.01, 0))
@
\caption{UCB admission with mosaic display.}
\label{fig:UCB-mosaic1}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic2,fig=TRUE,echo=FALSE,height=6,width=8>>=
mygp.max <- function(x,y) gp.max(x, y, level = (1 - c(0.1, 0.01)/6))
tabplot(margin.table(ucb, c(2,1,3)), panel = function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE, labels = FALSE, gp = mygp.max, space = 0.02, ...))
@
\caption{UCB admission with conditioned mosaic display.}
\label{fig:UCB-mosaic2}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<UCB-assoc,fig=TRUE,echo=FALSE,height=8,width=8>>=
tabplot(margin.table(ucb, c(2,1,3)), gp = mygp.max)
@
\caption{UCB admission with conditioned association plot.}
\label{fig:UCB-assoc}
\end{center}
\end{figure}

\bibliography{vcd}

\end{document}
